---
title: "Installing the unpak data snapshot"
author: "A Strand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing the unpak data snapshot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The unpakathon package is primarily a datafreeze of the unpak database.  It also provides some functions to help work with these data.  

This vignette shows how to set up the environment

#Prepping your computer

This package is designed to work with any modern operating system, but has not been tested on very many (linux and windows 8).  Without a doubt, to use the package on any system, you will need to:

 1. Install R on your computer (go to http://www.r-project.org).  If you are an old hand and use R to do everything including cook your breakfast in the morning, you still might want to install the latest version (as of 6/14/16 that's version 3.3.0 (Supposedly Educational))
 2. Install rstudio desktop on your computer as well (go to https://www.rstudio.com/products/RStudio/#Desktop).  Same admonitions from above apply.  The latest version is: 0.99.902
 3. Oh yeah, for any of this to work, you will need to have a network connection.
 
 
#Installing the data package

The package is available through github and not on CRAN. What this means is that you will need to install from github.com rather than using the normal R package installation protocol.  To do that, however, you will have to install the remarkable package, "devtools".  Here's how:

Start Rstudio and at the Rstudio console type:
```{r,eval=FALSE}
install.packages("devtools")
```

Once that completes, you can get crackin on the installation of our data on your computer:
```{r,eval=F}
library(devtools)
install_github("stranda/adjustPhenotypes") #this package needs to be installed first
install_github("stranda/unpakathon",force=T,build_vignettes=T)
```
These two "install_github()" commands will install a supporting package that I wrote (adjustPhenotypes) and then installs the main package called, "unpakathon".  This is going to take a while because unpakathon depends on lots of other packages which need to be installed.  They should be installed automatically, though.  When it finally ends, you should be able to say:
```{r,eval=F}
library(unpakathon)
browseVignettes("unpakathon")
```
If you see a web-page with different "help documents" you were successful.  The IntroUnpakData vignette is a good place to start


 
