---
title: "January 2016 unpakathon results"
author: "A Strand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summary of the work done during January meeting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

#Summary of plots we looked at during this initial hacakthon

#quick scan for weird data points



```{r,echo=F,results='hide'} 
#Load some libraries
library(unpakathonJan2016)
library(dplyr)
library(ggplot2)
library(reshape)
library(GGally)
```

#phenotypic distributions in the database

Here is the distribution of phenotypes across non-farm experiments.  The numbers correspond to the number of lines assessed for that phenotype *not* the number of plants

```{r}
with(phenolong %>% filter(meta.experiment != "farm") %>% select(accession,experiment,variable) %>% distinct(),
     table(variable,experiment))
```



Let's check the distribution of values for each phenotype

```{r,fig.height=13,fig.width=10}

ggplot(phenolong,aes(value)) + geom_histogram() + facet_wrap(~variable,scales = "free",ncol=4)

```

Several weird values for several phenotypes

Now look at problematic data (extreme values)

Here are the treatments and experiments with leaf.numbers greater than 50

```{r,eval=TRUE}
phenolong %>% filter(variable=="leaf.number",value>50)%>%select(experiment,treatment)%>%distinct()
```

Looks ok, these are all from cofc-second (big pots and fert), we'll keep em.

#Some automated vetting of observations
I'm going to flag points that are outside the central 90% of all values for a phenotype as suspect.  We can then switch this off and on and look at the outlier effect in analyses below
```{r}

phenoquant <- phenolong %>% group_by(variable) %>% summarise(quant05 = quantile(value,0.05,na.rm=T),quant95 = quantile(value,0.95,na.rm=T))
phenolong.quant = left_join(phenolong,phenoquant)
phenolong.quant$inrange <- with(phenolong.quant,ifelse((quant95>=value)&(quant05<=value),TRUE,FALSE))
phenolong = phenolong.quant
```

For example, look at the ggplot from above now
```{r,fig.height=13,fig.width=10}
ggplot(phenolong%>%filter(inrange),aes(value)) + geom_histogram() + facet_wrap(~variable,scales = "free",ncol=4)
```



#Biology instead of data integrity
##experiment 1 results
###Line means with columbia indicated in red and phytometers in blue.  Outliers excluded

```{r}
exp1long <- phenolong %>% filter(meta.experiment=="1",inrange) %>% group_by(variable,accession) %>% summarise(value=mean(value,na.rm=T)) 
exp1.col = exp1long %>% filter(accession%in%c("CS70000","CS60000"))
exp1.phyt = exp1long %>% filter(grepl("CS",accession))
phenos=unique(exp1long$variable)
tmp=sapply(phenos,function(ph){
  tmpphen = exp1long[exp1long$variable==ph,]
  if (ph=="fruitnum") {tmpphen = tmpphen[tmpphen$value<400,]}
  hist(tmpphen$value,main=paste("exp1",ph))
  tmpphyt=exp1.phyt %>% filter(variable==ph)
  tmpcol=exp1.col %>% filter(variable==ph)
  points(y=rep(0,dim(tmpphyt)[1]),x=tmpphyt$value, col="blue",pch=16, cex=1.1) #phytometers
  points(y=rep(0,dim(tmpcol)[1]),x=tmpcol$value, col="red",pch=16, cex=1.1) # columbia especially
  })

```

###Pairwise comparisons (outliers excluded)

```{r,fig.width=13,fig.height=14}
#phenos=c(unique(exp1long$variable))
phenos<- c("aborted.fruits","avg.fruit.length","basal.branch","days.to.bolt","diameter.at.bolt","fruitnum","germinants.14d","inflorescence.height", "ttl.maininfl.branch")
exp1long <- exp1long %>% filter(variable %in% phenos)
exp1wide <- cast(exp1long)
paircompare <- exp1wide[,-1]
paircompare=paircompare[complete.cases(paircompare),]
ggpairs(paircompare,lower=list(continuous="smooth"))
```

