---
title: "Outliers in the unpak data"
author: "A Strand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting outliers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

#Outliers

unpakathon provides some functions to help detect outliers for any particular variable.

First load the package
```{r,message=F,warning=F}
library(unpakathon)
```

then make a long-format dataframe containing the identities of fruitnum outliers.  Outliers are defined here as the top and bottom 1% of observations.  
```{r}
tmp = findOutlier(phenolong,"fruitnum",q=0.01,z.score=F)
head(tmp)
```
So the first 6 rows come from experiments cofc-1 and especially cofc-2 (the originals).  There are a total of 
```{r}
dim(tmp)[1]
```
individual measurements identified

##adjusting for differences among chambers

The 'z.score' value in the findOutlier function converts the phenotypes into numbers scaled by the sd in each growth chamber and then mean-subtracted.

```{r}
tmp = findOutlier(phenolong,"fruitnum",q=0.01,z.score=T)
head(tmp)
```
In this case there are a lot of experiment 2 treatment A outliers that are appearing.  There are a total of 
```{r}
dim(tmp)[1]
```
individual measurements identified by this technique (this should not really have changed much because the data are the same as above and we are using quantiles).

##Status of outliers and database.  
I'm planning on taking all the outlier information that April sent to me and do one hand-edit of the current database.  At this point, I think that will take a day, but I'd like to wait until I have all the experiment 2 data

##plotting the outliers with the rest of the data for a trait/experiment combination

The following million figures are all distributions of data so far with the findOutlier function and a cutoff of 0.1% point highlighted.

###One more thing:
The number by each point is the plantID from the builtin phenolong and phenowide datasets.  To look at a particular plantID, you can say (plantID 21935 has a --too-- extreme fruitnum):
```{r}
phenolong%>%filter(plantID=="21935")
```
to get the long form, or 
```{r}
phenowide%>%filter(plantID=="21935")
```
to get the wide form.  

A dataframe with just these outliers is available in the builtin R dataframe 'outliers' It's also in the 'dataclean' subdirectory of the unpackathon intall on your computer (look in the R library hierarchy).  You could look at that dataframe like this:
```{r}
head(outliers)
```


###ok here we go

```{r, echo=F,out.width=690,fig.width=6,fig.height=6}
library(ggplot2)
library(ggrepel)
cutoff=0.001
z.score=T
outliers=NULL
for (pheno in unique(phenolong$variable))
  for (trt in unique(phenolong$treatment[phenolong$variable==pheno]))
  {
#    pheno = "diameter.at.bolt"
#    trt="C"
    #print(pheno)
    #print(trt)
    
    tmpdf = phenolong %>% filter(variable %in% pheno) %>% filter(treatment %in% trt) %>%
      filter(meta.experiment %in% c("1","2","3")) 
    if (!pheno%in%c("germinated","branch.basalbranch"))
      if ((dim(tmpdf)[1]>5)&(var(tmpdf$value,na.rm=T)>0))
      {
        if (z.score)
        {
          tmpdf=scalePhenos(classifier=c("experiment","facility","treatment"),lineid="accession",dat=tmpdf)
          tmpdf$value[!is.finite(tmpdf$value)] = NA
          tmpdf$value=tmpdf$value - mean(tmpdf$value,na.rm=T)
        }
        tmpout=unique(findOutlier(tmpdf,pheno,cutoff))
        if (dim(tmpout)[1]>1)
        {
          tmpdf=unique(merge(tmpdf,tmpout,all=T))
          tmpdf$direction[is.na(tmpdf$direction)]="norm"
          outliers = rbind(outliers,tmpout)
          p = ggplot(tmpdf,aes(experiment,value)) + 
            geom_boxplot(outlier.size=0) +
            geom_point(data=tmpout,aes(colour=direction)) +
            geom_text_repel(data=tmpout,aes(label=plantID)) +
            ylab(paste0(pheno,ifelse(z.score==T," (scaled and centered)",""))) + 
            ggtitle(paste0("treatment -> ",trt,";   phenotype ->",pheno)) +
            theme(axis.text.x = element_text(angle = 90))
          print(p)
        }
      }
  }
```
```{r,include=F}
save(file="../data/outliers.rda",outliers)
write.csv(file="../inst/dataclean/outliers.csv",row.names=F,outliers)
```

