---
title: "How data are organized into strata"
author: "A Strand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How data are organized in to strata in the unpak database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


#Three levels of experimental organization

Each experiment unpak conducts is given a different experiment name, '1a1'.  These can be nested within a meta.experiment name (such as '1' for the experiment above).  In addition it is possible to specify a block within each experiment.  This provides two levels of hierarchy in the experiments, though these levels might not always be treated the same.


```{r include=F}
library(igraph)
library(unpakathon)
```


The following figure shows the relationship among meta experiments and the sub-components.  Many of the sub-components have blocking which though not shown here can be accessed in builtin R dataframes as the column 'block'
```{r,echo=F,fig.width=6,fig.height=8,out.width=600,out.height=800}
map=unique(phenolong[,c("experiment","meta.experiment")])
map=map[map[,2]!="farm",]
plot(graph_from_data_frame(map))
```

